R Notebook

### 2018-02-02 paper/R/04_Modelresults.Rmd
add l1norm, cosine similarity

After reviewing a sample of articles assigned to the respective topic, we assign a shorter name to the topics based on the most common words in order to improve readability and traceability.
```{r}
sagelabs$cov.betas[[1]]$problabels[,1:6]
```

@@

```{r}
k <- stmOut$settings$dim$K
y <- length(stmOut$settings$covariates$yvarlevels)
```

```{r}
## Find all combinations of ylevel
comb.matrix <- matrix(0, 
                      nrow= y^2, 
                      ncol = 2)

z <- 1
for (i in 1:y) {
  for (j in 1:y) {
    comb.matrix[z,1] <- stmOut$settings$covariates$yvarlevels[i]
    comb.matrix[z,2] <- stmOut$settings$covariates$yvarlevels[j]
    z <- z+1
  }
}
```

@@

# Populate matrix with JSD Values
x <- 1
for (i in 1:y) {

    #print(paste("ylevel = ", i))
    base <- exp(stmOut$beta$logbeta[[i]])
    compare1 <- exp(stmOut$beta$logbeta[[1]])
    compare2 <- exp(stmOut$beta$logbeta[[2]])
    # compare3 <- exp(stmOut$beta$logbeta[[3]])
    # compare4 <- exp(stmOut$beta$logbeta[[4]])
    # compare5 <- exp(stmOut$beta$logbeta[[5]])
    # compare6 <- exp(stmOut$beta$logbeta[[6]])

@@

    #print(paste("topic = ", j))
    kld.matrix[x,j] <- KLD(base[j,],compare1[j,])
    kld.matrix[x+1,j] <- KLD(base[j,],compare2[j,])
    # kld.matrix[x+2,j] <- JSD(base[j,],compare3[j,])
    # kld.matrix[x+3,j] <- JSD(base[j,],compare4[j,])
    # kld.matrix[x+4,j] <- JSD(base[j,],compare5[j,])
    # kld.matrix[x+5,j] <- JSD(base[j,],compare6[j,])

  }
    x <- x+y
}
```

@@

kld.df <- kld.df %>% mutate(News1 = comb.matrix[,1],
         News2 = comb.matrix[,2]) %>%
  melt(.,id=c("News1","News2"), value.name="KLD",
       variable.name = "topic") %>%
  mutate(topic = as.integer(topic)) %>%
  left_join(., topics.df, by="topic")
```

@@

    #print(paste("topic = ", j))
    jsd.matrix[x,j] <- JSD(base[j,],compare1[j,])
    jsd.matrix[x+1,j] <- JSD(base[j,],compare2[j,])
    # kld.matrix[x+2,j] <- JSD(base[j,],compare3[j,])
    # kld.matrix[x+3,j] <- JSD(base[j,],compare4[j,])
    # kld.matrix[x+4,j] <- JSD(base[j,],compare5[j,])
    # kld.matrix[x+5,j] <- JSD(base[j,],compare6[j,])

  }
    x <- x+y
}
```

@@

distance.df <- jsd.df %>% mutate(News1 = comb.matrix[,1],
         News2 = comb.matrix[,2]) %>% 
  melt(.,id=c("News1","News2"), value.name="JSD",
       variable.name = "topic") %>%
  mutate(topic = as.integer(topic)) %>%
  left_join(., kld.df, by=c("News1","News2","topic")) 
```

@@

p1 <- distance.df %>%
  filter(!topic %in% c("11","15")) %>%
  filter(News1!=News2) %>%
  ggplot(aes(News1, reorder(factor(topic_name),JSD), fill = KLD)) + 
  geom_tile() +
  #facet_wrap(~variable) + 
  scale_fill_gradientn(colours = cols) +
  labs(x="", y="") +
  theme_bw(base_family = "Roboto", base_size = 11) +
  theme(axis.text.y = element_blank(),
        legend.position = "top",
    #axis.text.x = element_text(angle=90),
        plot.background = element_rect(fill = 'transparent', colour = 'transparent'),
        strip.background = element_rect(fill = 'transparent', colour = 'transparent'),
        axis.ticks = element_line(colour = 'transparent'))

p2 <- distance.df %>%
  filter(!topic %in% c("11","15")) %>%
  filter(News1!=News2) %>%
  filter(News1=="private") %>%
  ggplot(aes(News1, reorder(factor(topic_name),JSD), fill = JSD)) + 
  geom_tile() +
  #facet_wrap(~variable) + 
  scale_fill_gradientn(colours = cols) +
  labs(x="", y="") +
  theme_bw(base_family = "Roboto", base_size = 11) +
  theme(axis.text.x = element_blank(),
        legend.position = "top",
    #axis.text.x = element_text(angle=90),
        plot.background = element_rect(fill = 'transparent', colour = 'transparent'),
        strip.background = element_rect(fill = 'transparent', colour = 'transparent'),
        axis.ticks = element_line(colour = 'transparent'))

p2+p1

ggsave(filename = "../figs/distance.png", device = "png",width = 10, height = 8,
dpi = 600)
```

>>>>

After reviewing a sample of articles assigned to the respective topic, we assign a shorter name to the topics based on the most common words in order to improve readability and traceability.
```{r}
sagelabs$cov.betas[[2]]$problabels[,1:6]
```

@@

```{r}
k <- stmOut$settings$dim$K
p <- length(stmOut$settings$covariates$yvarlevels)
```

```{r}
## Find all combinations of ylevel
comb.matrix <- matrix(0, 
                      nrow= p^2, 
                      ncol = 2)

z <- 1
for (i in 1:p) {
  for (j in 1:p) {
    comb.matrix[z,1] <- stmOut$settings$covariates$yvarlevels[i]
    comb.matrix[z,2] <- stmOut$settings$covariates$yvarlevels[j]
    z <- z+1
  }
}
```

### Calculate L1 Norm
```{r}
# Create empty matrix
l1.matrix <- matrix(0, nrow = nrow(comb.matrix), ncol = k)
names(l1.matrix) <- 1:k

# Populate matrix with JSD Values
z <- 1
for (i in 1:p) {

    #print(paste("ylevel = ", i))
    base <- exp(stmOut$beta$logbeta[[i]])
    compare1 <- exp(stmOut$beta$logbeta[[1]])
    compare2 <- exp(stmOut$beta$logbeta[[2]])

    #print(paste("topic = ", j))
    l1.matrix[z,j] <- sum(abs(base[j,]-compare1[j,]))
    l1.matrix[z+1,j] <- sum(abs(base[j,]-compare2[j,]))

  }
    z <- z+p
}

```{r}
## Create DF
l1.df <- as.data.frame(l1.matrix) 
colnames(l1.df) <- 1:k

l1.df <- l1.df %>% 
  mutate(News1 = comb.matrix[,1],
         News2 = comb.matrix[,2]) %>%
  melt(.,id=c("News1","News2"), value.name="L1Norm",
       variable.name = "topic") %>%
  mutate(topic = as.integer(topic)) %>%
  left_join(., topics.df, by="topic")
```

### Calculate Cosine Similarity
```{r}
# Create empty matrix
cosine.matrix <- matrix(0, nrow = nrow(comb.matrix), ncol = k)
names(cosine.matrix) <- 1:k

# Populate matrix with JSD Values
z <- 1
for (i in 1:p) {

    #print(paste("ylevel = ", i))
    base <- exp(stmOut$beta$logbeta[[i]])
    compare1 <- exp(stmOut$beta$logbeta[[1]])
    compare2 <- exp(stmOut$beta$logbeta[[2]])

    #print(paste("topic = ", j))
    cosine.matrix[z,j] <- CosSim(base[j,],compare1[j,])
    cosine.matrix[z+1,j] <- CosSim(base[j,],compare2[j,])

  }
    z <- z+p
}

```{r}
## Create DF
cosine.df <- as.data.frame(cosine.matrix) 
colnames(cosine.df) <- 1:k

cosine.df <- cosine.df %>% 
  mutate(News1 = comb.matrix[,1],
         News2 = comb.matrix[,2]) %>%
  melt(.,id=c("News1","News2"), value.name="CosineSimilarity",
       variable.name = "topic") %>%
  mutate(topic = as.integer(topic)) %>%
  left_join(., l1.df, by=c("News1","News2","topic"))
```

@@

# Populate matrix with JSD Values
z <- 1
for (i in 1:p) {

    #print(paste("ylevel = ", i))
    base <- exp(stmOut$beta$logbeta[[i]])
    compare1 <- exp(stmOut$beta$logbeta[[1]])
    compare2 <- exp(stmOut$beta$logbeta[[2]])

@@

    #print(paste("topic = ", j))
    kld.matrix[z,j] <- KLD(base[j,],compare1[j,])
    kld.matrix[z+1,j] <- KLD(base[j,],compare2[j,])

  }
    z <- z+p
}
```

@@

kld.df <- kld.df %>% mutate(News1 = comb.matrix[,1],
         News2 = comb.matrix[,2]) %>%
  melt(.,id=c("News1","News2"), value.name="KLD",
       variable.name = "topic") %>%
  mutate(topic = as.integer(topic)) %>%
  left_join(., cosine.df, by=c("News1","News2","topic"))
```

@@

    #print(paste("topic = ", j))
    jsd.matrix[z,j] <- JSD(base[j,],compare1[j,])
    jsd.matrix[z+1,j] <- JSD(base[j,],compare2[j,])

  }
    z <- z+p
}
```

@@

distance.df <- jsd.df %>% mutate(News1 = comb.matrix[,1],
         News2 = comb.matrix[,2]) %>% 
  melt(.,id=c("News1","News2"), value.name="JSD",
       variable.name = "topic") %>%
  mutate(topic = as.integer(topic)) %>%
  left_join(., kld.df, by=c("News1","News2","topic"))

rm(l1.df, jsd.df, cosine.df, kld.df)
```

@@

p1 <- distance.df %>%
  filter(!topic %in% c("11","15")) %>%
  filter(News1!=News2) %>%
  filter(News1=="private") %>%
  ggplot(aes(News1, reorder(factor(topic_name),JSD), fill = JSD)) + 
  geom_tile() +
  scale_fill_gradientn(colours = cols) +
  labs(x="", y="") +
  theme_bw(base_family = "Roboto", base_size = 11) +
  theme(axis.text.x = element_blank(),
        legend.position = "top",
        plot.background = element_rect(fill = 'transparent', colour = 'transparent'),
        strip.background = element_rect(fill = 'transparent', colour = 'transparent'),
        axis.ticks = element_line(colour = 'transparent'))

p2 <- distance.df %>%
  filter(!topic %in% c("11","15")) %>%
  filter(News1!=News2) %>%
  ggplot(aes(News1, reorder(factor(topic_name),JSD), fill = KLD)) + 
  geom_tile() +
  scale_fill_gradientn(colours = cols) +
  labs(x="", y="") +
  theme_bw(base_family = "Roboto", base_size = 11) +
  theme(axis.text.y = element_blank(),
        legend.position = "top",
        plot.background = element_rect(fill = 'transparent', colour = 'transparent'),
        strip.background = element_rect(fill = 'transparent', colour = 'transparent'),
        axis.ticks = element_line(colour = 'transparent'))

p3 <- distance.df %>%
  filter(!topic %in% c("11","15")) %>%
  filter(News1!=News2) %>%
  filter(News1=="private") %>%
  ggplot(aes(News1, reorder(factor(topic_name),JSD), 
             fill = CosineSimilarity)) + 
  geom_tile() +
  scale_fill_gradientn(colours = cols) +
  labs(x="", y="") +
  theme_bw(base_family = "Roboto", base_size = 11) +
  theme(axis.text = element_blank(),
        legend.position = "top",
        plot.background = element_rect(fill = 'transparent', colour = 'transparent'),
        strip.background = element_rect(fill = 'transparent', colour = 'transparent'),
        axis.ticks = element_line(colour = 'transparent'))

p4 <- distance.df %>%
  filter(!topic %in% c("11","15")) %>%
  filter(News1!=News2) %>%
  filter(News1=="private") %>%
  ggplot(aes(News1, reorder(factor(topic_name),JSD), 
             fill = L1Norm)) + 
  geom_tile() +
  scale_fill_gradientn(colours = cols) +
  labs(x="", y="") +
  theme_bw(base_family = "Roboto", base_size = 11) +
  theme(axis.text = element_blank(),
        legend.position = "top",
        plot.background = element_rect(fill = 'transparent', colour = 'transparent'),
        strip.background = element_rect(fill = 'transparent', colour = 'transparent'),
        axis.ticks = element_line(colour = 'transparent'))

(p1 | p2 | p3 | p4 )

ggsave(filename = "../figs/distance.png", device = "png",width = 12, height = 8,
dpi = 600)
```

### 2018-02-02 paper/doc/structure.tex
add l1norm, cosine similarity

@@

where the word probability $\beta_{dkv}$ is parameterized in terms of log-transformed rate deviations from the rates of a corpus-wide background distribution $m_v$. The log-transformed rate deviations can then be specified by a collection of parameters $\lbrace \boldsymbol{\kappa} \rbrace$, where $\kappa^{(t)}$ is a $K$-by-$V$ matrix containing the log-transformed rate deviations for each topic $k$ and term $v$, over the baseline log-transformed rate for term $v$. This matrix is the same for all $A$ levels of covariates. To put it differently, $\kappa^{(t)}$ indicates the importance of the term $v$ given topic $k$ regardless of the covariates. Similarly, $\kappa^{(c)}$ is a $A$-by-$V$ matrix, indicating the importance of the term $v$ given the covariate level $k$ regardless of the topic. Finally, $\kappa^{(i)}$ is a $A$-by-$K$-by-$V$ matrix, collecting the covariate-topic effects.

@@

The STM maximizes the posterior likelihood that the observed data were generated by the above data-generating process using an iterative approximation-based variational expectation-maximization algorithm\footnote{A technical description of this maximization process can be found in \citet{roberts_model_2016}} available in R's stm package \citep{roberts_stm:_2016}. The process gives us two posterior distribution parameter: (1) $\beta$ is a $K$-by-$V$ matrix (where $K=$ number of topics and $V=$ vocabulary), where the entry $\beta_{kv}$ can be interpreted as the probability of observing the $v$-th word in topic $k$. As we let covariate levels (the news provider) effect the word-distribution of a topic, we will get as many $\beta$ matrices as news provider. (2) $\theta$ is a $D$-by-$V$ matrix of the document-topic distributions, where the entry $\theta_{dk}$ can be interpreted as the proportion of words in document $d$ which arise from topic $k$, or rather as the probability that document $d$ deals about topic $k$. We will use these probability distributions to compare the content of public and private news provider in Section \ref{ch_empirical}.

@@

This process revealed that a model with 40 topics best reflects the structure in the corpus. Furthermore, we use the source of each article (the media house) and the month it was published as covariates in the topic prevalence. In other words, the probability distribution of topics depends on the editorial strategy of a media house as well as on the month the article was published. We also include news agency as a covariate affecting topical content to estimate how topics are discussed in different ways by different news agencies. To address problems due to non-convexity, we rely on the spectral initialization approach advocated by \citet{roberts_navigating_2016}.

@@

@@

@@

for each news provider we get a respective $\beta$, $K$x$V$ matrix. We then use each row (corresponding to each topic) to calculate the similarity of the word-topic distribution between the news provider using the square root of the Jensen-Shannon (JS) Divergence.

\todo{Compare different metrices}
L1 distance, cosine similarity metric (essentially the dot product rescaled by the L2 norm of the vectors).

From the STM model we get a $KxV$ matrix (where $K=$number of topics and $V=$vocabulary) of the whole corpus that contains the probability of seeing each word of the vocabulary $V$ within topic $k$ conditional on the news provider. That is, for every news provider we get a $KxV$ matrix. We use each row (corresponding to each topic) to calculate the similarity of the word-topic distribution between the news provider using the square root of the Jensen-Shannon (JS) Divergence.

A low JSD value indicates stronger similarities.

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.8\textwidth,keepaspectratio]{../figs/distance}
		\caption{Similarity measures of word-topic probabilities}
	\end{center}
\end{figure}

>>>>

\subsection{Similar Research}

M.Paul (Cross-Collection Topic Models: Automatically Comparing and Contrasting Text): compares editorial differences between media sources, using cross-collection latent Dirichlet allocation (ccLDA), an LDA-based approach that incorporates differences in document metadata. They use a dataset of 623 news articles from August 2008 from two American media outlets - msnbc.com and foxnews.com - to analyse how they discuss topics differently. Comparing the top words of the word-topic distribution, they find some content differences between the two.

@@

where the word probability $\beta_{dkv}$ is parameterized in terms of log-transformed rate deviations from the rates of a corpus-wide background distribution $m_v$. The log-transformed rate deviations can then be specified by a collection of parameters $\lbrace \boldsymbol{\kappa} \rbrace$, where $\kappa^{(t)}$ is a $K$-by-$V$ matrix containing the log-transformed rate deviations for each topic $k$ and term $v$, over the baseline log-transformed rate for term $v$. This matrix is the same for all $A$ levels of covariates. To put it differently, $\kappa^{(t)}$ indicates the importance of the term $v$ given topic $k$ regardless of the covariates. Similarly, $\kappa^{(c)}$ is a $A$-by-$V$ matrix, indicating the importance of the term $v$ given the covariate level $c$ regardless of the topic. Finally, $\kappa^{(i)}$ is a $A$-by-$K$-by-$V$ matrix, collecting the covariate-topic effects.

@@

The STM maximizes the posterior likelihood that the observed data were generated by the above data-generating process using an iterative approximation-based variational expectation-maximization algorithm\footnote{A technical description of this maximization process can be found in \citet{roberts_model_2016}} available in R's stm package \citep{roberts_stm:_2016}. The process gives us two posterior distribution parameter: (1) $\beta$ is a $K$-by-$V$ matrix (where $K=$ number of topics and $V=$ vocabulary), where the entry $\beta_{kvc}$ can be interpreted as the probability of observing the $v$-th word in topic $k$ for the covariate level $c$. (2) $\theta$ is a $D$-by-$V$ matrix of the document-topic distributions, where the entry $\theta_{dk}$ can be interpreted as the proportion of words in document $d$ which arise from topic $k$, or rather as the probability that document $d$ deals about topic $k$. We will use these probability distributions to compare the content of public and private news provider in Section \ref{ch_empirical}.

@@

This process revealed that a model with 40 topics best reflects the structure in the corpus. Furthermore, we use the ownership of the news provider of each article (public or private) and the month it was published as covariates in the topic prevalence. In other words, the probability distribution of topics depends on the business-model of a media house as well as on the month the article was published. We also include ownership as a covariate affecting topical content to estimate how topics are discussed in different ways by different news agencies. To address problems due to non-convexity, we rely on the spectral initialization approach advocated by \citet{roberts_navigating_2016}.

@@

In this section we summarize the results of the STM and apply different measures to analyze the content differences between public and private ownership.

@@

\citep{roberts_navigating_2016}: Poliblog corpus to examine the role of partisanship in topical coverage.

@@

Since the word distribution of a topic is a simple mapping of text in space, we can use their word distribution $\beta$ to compute the similarity. More precisely, we get a respective $\beta$, $K$x$V$ matrix for each covariate level. We use each row (corresponding to each topic) to calculate the similarity of the word-topic distribution between the news provider using the square root of the Jensen-Shannon (JS) Divergence. Jensen-Shannon Divergence is a positive definite measure, satisfying the following conditions: $D_{js}(a,b) \geq 0$, $D_{js}(a,b)=0$ iff $(a=b)$. It is also symmetric: $D_{js}(a,b)=D_{js}(b,a)$. Let $p(x)$ and $q(x)$ be two probability density functions, the Jensen-Shannon distance $D_{js}(p,q)$ between this two distributions is defined as:

\begin{align*}
	D_{js}(p,q)=\frac{1}{2}D_{KL}(p,\frac{p+q}{2})+\frac{1}{2}D_{KL}(q,\frac{p+q}{2})
\end{align*}

where $D_{KL}$ is the Kullback-Leibler divergence between $p$ and $q$ defined as:

\begin{align*}
	D_{KL}(p,q)=\sum_{i=1}^T p_i ln \frac{p_i}{q_i}
\end{align*}

So $D_{js} \to 0$ indicates stronger similarities. Since $D_{js}$ does not satisfy the triangular inequality condition $D_{js}(a,c)\leq D_{js}(a,c)+D_{js}(b,c)$, the JSD is not considered to be a real metric. However, we can use the square root of JSD as a real distance metric \citep{endres_new_2003}. We will compare this metric with a set of other similarity measures recently used in the literature to measure the difference between word-topic distributions.

\citep{roberts_navigating_2016}: They use the $L_1$ norm of the word-topic distributions to compare topics from different models. The $L_1$ norm is the sum of the absolute value of the difference between two vectors. Its defined as

\begin{align*}
	L_1=\sum_v |\beta_{kvi}-\beta_{kvj}|
\end{align*}

where $i$ and $j$ indicate different covariate levels. This measure has a range of $[0,2]$, where $ L_1 \rightarrow 0$ indicates strong similarities. They compare this measure with a cosine similarity metric, which is essentially the dot product rescaled by the $L_2$ norm of the vectors and is defined as

\begin{align*}
	\text{cos}(\theta)=\frac{\sum^d_{v=1}\beta_{kvi}\beta{kvj}}{\sqrt{\sum^d_{v=1}\beta_{kvi}^2}\sum^d_{v=1}\beta_{kvj}^2}
\end{align*}

Again, $i, j$ indicate different covariate levels. However, different to the above mentioned metrics the cosine similarity has a range of $[0,1]$ and $cos(\theta) \to 0$ indicate less similarity.

Figure \ref{fig_distance} displays the different distance measures

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.8\textwidth,keepaspectratio]{../figs/distance}
		\caption{Similarity measures of word-topic probabilities}
		\label{fig_distance}
	\end{center}
\end{figure}

\todo{check topics that are most different}
